\ttitle

### 2015-08-21 Chapters/CPET_Outcomes.tex
Review 1 - CPET_Outcomes
- Complete

\subsection{Aims}
The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative adverse events as well as fitness for adjuvant therapy in patients undergoing major pancreatic surgery.

@@

\section{Patients and Methods}
Patients who underwent pancreaticoduodenectomy or total pancreatectomy for pancreatic head lesions between August 2008, when cardiopulmonary exercise testing was first used for fitness assessment at our hospital, and January 2012 were considered for this retrospective study. 
Patients who had not undergone cardiopulmonary exercise testing as part of their preoperative assessment and patients who underwent cardiopulmonary exercise testing but did not undergo surgery were excluded.

Data on patient demographics, comorbidity including cardiovascular and respiratory disease, preoperative blood tests, chest x-ray and cardiopulmonary exercise tests were collected from prospectively maintained databases (march 2009 - January 2012) and case note review (August 2008 - March 2009). 
Data was also collected for patients who did not undergo cardiopulmonary exercise testing to allow comparison with the study group. 
The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease including hypertension, ischaemic heart disease and heart failure, respiratory disease causing breathlessness on exertion and COPD, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) as described previously.

Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
An electrically-braked cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. 
The test was stopped at maximum exercise tolerance, significant ischaemic changes on ECG or for other safety reasons. 
The VO$_2$AT was calculated using the V-slope \parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents \parencite{sue_metabolic_1988} methods. 
Low VO$_2$AT was defined as oxygen consumption less than 10ml/kg/min based on work by Snowden and co-workers \parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.

The decision to operate was based on overall preoperative evaluation of the patient’s comorbid conditions and performance status and not exclusively on the result of cardiopulmonary exercise testing. 
Whilst the results of cardiopulmonary exercise tests were available to the clinicians before surgery, no specific changes were made to perioperative management based exclusively on these results. 
These results were used in conjunction with other established forms of preoperative evaluation for risk assessment and perioperative care. 
All patients were routinely admitted to the surgical high dependency unit unless intra-operative events or postoperative complications required admission to the intensive care unit. 
Patients were discharged after resolution of organ dysfunction and/or sepsis and when nutrition, analgesia and mobilisation were adequately established to the clinician's and patient's satisfaction.

Postoperative adverse events were recorded using internationally recognised definitions. 
The International Study Group for Pancreatic Surgery (ISGPS) definitions were used to classify pancreatic fistulae \parencite{bassi_postoperative_2005} and post-operative haemorrhage \parencite{wente_postpancreatectomy_2007}. 
The Clavien-Dindo (CD) classification \parencite{clavien_clavien-dindo_2009, dindo_classification_2004} was used to grade other complications and CD grades III-V were considered major. 
Multiple admissions to critical care as well as re-operations were recorded. 
Operative mortality was defined as postoperative death in-hospital regardless of duration of stay or occurring within 30 days of the surgery. 
All complications were discussed at a weekly multidisciplinary meeting attended by three pancreatic surgeons and a radiologist with a specialist interest in pancreatic diseases and recorded in a prospective database.

@@

\section{Results}
One hundred and twenty-nine patients had undergone pancreaticoduodenectomy (n=127), sub-total pancreatectomy (n=1) or total pancreatectomy (n=1) during the study period. 
Sub-total and total pancreatectomy were performed in patients scheduled for a pancreaticoduodenectomy but were found to have pancreatic remnants either too friable or too atrophic during the operation to perform an anastomosis. 
Of these, 100 patients (pancreaticoduodenectomy - 98, sub-total/total pancreatectomy - 2) had undergone cardiopulmonary exercise testing as part of their preoperative assessment and were included in the study. 
Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intraductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=4) or chronic pancreatitis (n=2).

@@

The median VO$_2$AT was 10.3 ml/kg/min (inter-quartile range, IQR 8.8 - 11.6). 
The VO$_2$AT was less than 10ml/kg/min in 49 patients. 
The distribution of VO$_2$AT across the study cohort is shown in Figure \ref{fig:cpet_outcomes_dist_of_AT}.

@@

The relationship between VO$_2$AT and major postoperative adverse events including mortality is shown in Table \ref{table:cpet_outcomes_table2}. 
Patients with VO$_2$AT less than 10ml/kg/min had significantly greater incidence of postoperative pancreatic fistula (35.4\% vs.16\%, p=0.028) as well as major intra-abdominal abscesses (Clavien-Dindo Grade III - V, 22.4\% vs.7.8\%, p=0.042). 
While there was an association between low VO$_2$AT and grade of pancreatic fistula, this was not statistically significant (p=0.091). 
There was no association between low VO$_2$AT and cardiopulmonary complications or postoperative mortality. 
Major cardiopulmonary complications occurred more often in patients with major intra-abdominal adverse events including major intra-abdominal abscesses or Grade B and C pancreatic fistulae or haemorrhage than in patients who did not have these complications (5/31,16.1\% vs. 2/69,2.9\%, p=0.017). 
Postoperative mortality was not associated with VO$_2$AT (HR 0.77, 95\% CI 0.16-3.61, p 0.737) or the POSSUM Physiology Score (HR 0.39, 95\% CI 0.07-2.12, p 0.277). 
Postoperative mortality was associated with postoperative pancreatic fistula (n=5), post-pancreatectomy haemorrhage (n=3), major intra-abdominal sepsis (n=6) and major cardiorespiratory complications (n=4) with 6 patients requiring radiological or operative intervention.

@@

The median length of postoperative stay was 17 days (IQR 13 - 26). 
The median cumulative length of stay in critical care was 7 days (IQR 6 - 12). 
Twenty-six patients were admitted to critical care more than once. 
The relationship between preoperative clinico-pathological characteristics and length of postoperative stay in patients who were discharged from hospital (n=93) is shown in Table \ref{table:cpet_outcomes_table3}. 
On univariate analysis, age over 65 years (p=0.072) and low VO$_2$AT (p=0.010) were associated with prolonged postoperative stay. 
On multivariate Cox proportional hazards regression analysis, VO$_2$AT less than 10ml/kg/min (hazard ratio 1.74, 95\% confidence intervals 1.14-2.65, p=0.010) was the only significant factor associated with prolonged postoperative stay. 
A Kaplan-Meier plot for the probability of remaining in hospital over time for patients with low and normal VO$_2$ATs is shown in Figure \ref{fig:cpet_outcomes_km_at_los}. 
Patients with a low VO$_2$AT stayed a median 6 days longer in hospital (14 versus 20 days, Mann-Whitney Test p=0.001). 
There was no significant association between any of the preoperative factors including VO$_2$AT and length of critical care stay or number of critical care admissions.

@@

% Figure 2
\begin{figure}[h]
\centering
\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_km_at_los}
\includegraphics[width=1\linewidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
\caption{Kaplan-Meier Plot of postoperative length of stay in patients with VO$_2$AT $>=$ 10ml/kg/min versus $<$ 10ml/kg/min.}
\label{fig:cpet_outcomes_km_at_los}
\end{figure}

The relationship between clinico-pathological patient factors and receipt of adjuvant therapy is shown in Table \ref{table:cpet_outcomes_table4}. 
Fifty-five patients were included in the analysis. 
Patients were excluded if chemotherapy was not indicated (n=28), in the event of operative mortality (n=7), if chemotherapy was offered but declined by the patient (n=4), or where they had not been seen by an oncologist yet (n=6). 
On binary logistic regression analysis, VO$_2$AT less than 10ml/kg/min was the only preoperative factor that was associated with with non-receipt of adjuvant therapy (HR 6.30, 95\% CI 1.25-31.75, p=0.026).

@@

In the present study, VO$_2$AT was less than 10ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. 
The proportion of patients with VO$_2$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit. 
While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}. 
Older and co-workers reported in 1993 that low VO$_2$AT less than 11ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.

@@

The results of this study are consistent with the results of the study by Ausania and co-workers \parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min. 
However, this study did not report the association between VO$_2$AT and receipt of adjuvant therapy.

@@

This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of ‘prehabilitation’ \parencite{topp_effect_2002}. 
It has been suggested that prehabilitation not only improves aerobic capacity \parencite{jones_effects_2007} but may also improve postoperative recovery \parencite{mayo_impact_2011, pehlivan_effects_2011}. 
The results of this study show that impaired aerobic capacity is associated with postoperative adverse events. 
Therefore, it would appear that prehabilitation using interventions such as exercise and nutrition, by improving physiological fitness, may have a role in improving postoperative outcomes after major pancreatic surgery and may improve the proportion of patients receiving adjuvant therapy.

>>>>

\subsection{Aim}
The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative adverse events as well as fitness for adjuvant therapy in patients undergoing major pancreatic surgery.

@@

\section{Patients and Methods}

\subsection{Patients}
Patients who underwent pancreaticoduodenectomy or total pancreatectomy for pancreatic head lesions between August 2008, when cardiopulmonary exercise testing was first used for fitness assessment at our hospital, and January 2012 were considered for this retrospective study. 
Patients who had not undergone cardiopulmonary exercise testing as part of their preoperative assessment and patients who underwent cardiopulmonary exercise testing but did not undergo surgery were excluded.

\subsection{Preoperative data}
Data on patient demographics, comorbidity including cardiovascular and respiratory disease, preoperative blood tests, chest x-ray and cardiopulmonary exercise tests were collected from prospectively maintained databases (march 2009 - January 2012) and case note review (August 2008 - March 2009). 
Data was also collected for patients who did not undergo cardiopulmonary exercise testing to allow comparison with the study group. 
The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease including hypertension, ischaemic heart disease and heart failure, respiratory disease causing breathlessness on exertion and COPD, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) as described previously.

\subsection{Cardiopulmonary exercise test}
Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
An electrically-braked cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. 
The test was stopped at maximum exercise tolerance, significant ischaemic changes on ECG or for other safety reasons. 
The VO$_2$AT was calculated using the V-slope \parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents \parencite{sue_metabolic_1988} methods. 
Low VO$_2$AT was defined as oxygen consumption less than 10 ml/kg/min based on work by Snowden and co-workers \parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.
A detailed description of the cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.

\subsection{Perioperative care}
The decision to operate was based on overall preoperative evaluation of the patient's comorbid conditions and performance status and not exclusively on the result of cardiopulmonary exercise testing. 
Whilst the results of cardiopulmonary exercise tests were available to the clinicians before surgery, no specific changes were made to perioperative management based exclusively on these results. 
These results were used in conjunction with other established forms of preoperative evaluation for risk assessment and perioperative care. 
All patients were routinely admitted to the surgical high dependency unit unless intra-operative events or postoperative complications required admission to the intensive care unit. 
Patients were discharged after resolution of organ dysfunction and/or sepsis and when nutrition, analgesia and mobilisation were adequately established to the clinician's and patient's satisfaction.

\subsection{Outcome measures}
Postoperative adverse events were recorded using internationally recognised definitions. 
The International Study Group for Pancreatic Surgery (ISGPS) definitions were used to classify pancreatic fistulae \parencite{bassi_postoperative_2005} and post-operative haemorrhage \parencite{wente_postpancreatectomy_2007}. 
The Clavien-Dindo (CD) classification \parencite{clavien_clavien-dindo_2009, dindo_classification_2004} was used to grade other complications and CD grades III-V were considered major. 
Multiple admissions to critical care as well as re-operations were recorded. 
Operative mortality was defined as postoperative death in-hospital regardless of duration of stay or occurring within 30 days of the surgery. 
All complications were discussed at a weekly multidisciplinary meeting attended by three pancreatic surgeons and a radiologist with a specialist interest in pancreatic diseases and recorded in a prospective database.

@@

\section{Results}

\subsection{Clinico-pathological characteristics}
One hundred and twenty-nine patients had undergone pancreaticoduodenectomy (n=127), sub-total pancreatectomy (n=1) or total pancreatectomy (n=1) during the study period. 
Sub-total and total pancreatectomy were performed in patients scheduled for a pancreaticoduodenectomy but were found to have pancreatic remnants either too friable or too atrophic during the operation to perform an anastomosis. 
Of these, 100 patients (pancreaticoduodenectomy - 98, sub-total/total pancreatectomy - 2) had undergone cardiopulmonary exercise testing as part of their preoperative assessment and were included in the study. 
Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intra-ductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=4) or chronic pancreatitis (n=2).

@@

The median VO$_2$AT was 10.3 ml/kg/min (inter-quartile range, IQR 8.8 - 11.6). 
The VO$_2$AT was less than 10 ml/kg/min in 49 patients. 
The distribution of VO$_2$AT across the study cohort is shown in Figure \ref{fig:cpet_outcomes_dist_of_AT}.

@@

\subsection{Anaerobic threshold vs. complications}
The relationship between VO$_2$AT and major postoperative adverse events including mortality is shown in Table \ref{table:cpet_outcomes_table2}. 
Patients with VO$_2$AT less than 10 ml/kg/min had significantly greater incidence of postoperative pancreatic fistula (35.4\% vs.16\%, p=0.028) as well as major intra-abdominal abscesses (Clavien-Dindo Grade III - V, 22.4\% vs.7.8\%, p=0.042). 
While there was an association between low VO$_2$AT and grade of pancreatic fistula, this was not statistically significant (p=0.091). 
There was no association between low VO$_2$AT and cardiopulmonary complications or postoperative mortality. 
Major cardiopulmonary complications occurred more often in patients with major intra-abdominal adverse events including major intra-abdominal abscesses or Grade B and C pancreatic fistulae or haemorrhage than in patients who did not have these complications (5/31,16.1\% vs. 2/69,2.9\%, p=0.017). 
Postoperative mortality was not associated with VO$_2$AT (HR 0.77, 95\% CI 0.16-3.61, p 0.737) or the POSSUM Physiology Score (HR 0.39, 95\% CI 0.07-2.12, p 0.277). 
Postoperative mortality was associated with postoperative pancreatic fistula (n=5), post-pancreatectomy haemorrhage (n=3), major intra-abdominal sepsis (n=6) and major cardiorespiratory complications (n=4) with 6 patients requiring radiological or operative intervention.

@@

\subsection{Anaerobic threshold vs. length of stay}
The median length of postoperative stay was 17 days (IQR 13 - 26). 
The median cumulative length of stay in critical care was 7 days (IQR 6 - 12). 
Twenty-six patients were admitted to critical care more than once. 
The relationship between preoperative clinico-pathological characteristics and length of postoperative stay in patients who were discharged from hospital (n=93) is shown in Table \ref{table:cpet_outcomes_table3}. 
On univariate analysis, age over 65 years (p=0.072) and low VO$_2$AT (p=0.010) were associated with prolonged postoperative stay. 
On multivariate Cox proportional hazards regression analysis, VO$_2$AT less than 10 ml/kg/min (hazard ratio 1.74, 95\% confidence intervals 1.14-2.65, p=0.010) was the only significant factor associated with prolonged postoperative stay. 
A Kaplan-Meier plot for the probability of remaining in hospital over time for patients with low and normal VO$_2$ATs is shown in Figure \ref{fig:cpet_outcomes_km_at_los}. 
Patients with a low VO$_2$AT stayed a median 6 days longer in hospital (14 versus 20 days, Mann-Whitney Test p=0.001). 
There was no significant association between any of the preoperative factors including VO$_2$AT and length of critical care stay or number of critical care admissions.

@@

% Figure 2
\begin{figure}[h]
\centering
\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_km_at_los}
\includegraphics[width=1\linewidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
\caption{Kaplan-Meier Plot of postoperative length of stay in patients with VO$_2$AT $>=$ 10 ml/kg/min versus $<$ 10 ml/kg/min.}
\label{fig:cpet_outcomes_km_at_los}
\end{figure}

\subsection{Anaerobic threshold vs. receipt of adjuvant therapy}
The relationship between clinico-pathological patient factors and receipt of adjuvant therapy is shown in Table \ref{table:cpet_outcomes_table4}. 
Fifty-five patients were included in the analysis. 
Patients were excluded if chemotherapy was not indicated (n=28), in the event of operative mortality (n=7), if chemotherapy was offered but declined by the patient (n=4), or where they had not been seen by an oncologist yet (n=6). 
On binary logistic regression analysis, VO$_2$AT less than 10 ml/kg/min was the only preoperative factor that was associated with with non-receipt of adjuvant therapy (HR 6.30, 95\% CI 1.25-31.75, p=0.026).

@@

In the present study, VO$_2$AT was less than 10 ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. 
The proportion of patients with VO$_2$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit. 
While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}. 
Older and co-workers reported in 1993 that low VO$_2$AT less than 11 ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.

@@

The results of this study are consistent with the findings by Ausania and co-workers \parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min. 
However, this study did not report the association between VO$_2$AT and receipt of adjuvant therapy.

@@

This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of `prehabilitation' \parencite{topp_effect_2002}. 
It has been suggested that prehabilitation not only improves aerobic capacity \parencite{jones_effects_2007} but may also improve postoperative recovery \parencite{mayo_impact_2011, pehlivan_effects_2011}. 
The results of this study show that impaired aerobic capacity is associated with postoperative adverse events. 
Therefore, it would appear that prehabilitation using interventions such as exercise and nutrition, by improving physiological fitness, may have a role in improving postoperative outcomes after major pancreatic surgery and may improve the proportion of patients receiving adjuvant therapy.
